---
title: Stochastic Volatility Models
author: Ozgur Asar
date: '2019-12-21'
slug: stochastic-volatility-models
categories:
  - R
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2019-12-21T22:27:23+03:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

**Important note**: This page is only educational, i.e. do not 
take it too serious to make investments!

## MAVI Stock Data 

Below is the dynamic time-series plot of 
the daily closing MAVI stock data, between 15 June 2017 
and 20 December 2019; there are 656 days in total. 
Moving the cursor 
on the plot will tell the date and the 
value of the stock, and the plot can be zoomed 
by slicing it and un-zoomed by double left click.

\vspace{5in}

```{r echo = FALSE}
suppressMessages(library(tidyverse))
suppressMessages(library(quantmod))
suppressMessages(library(dygraphs))
options("getSymbols.warning4.0" = FALSE)
dat <- getSymbols("MAVI.IS", auto.assign = FALSE, warnings = FALSE)

dat %>% Cl() %>% dygraph()
```




Corresponding one-period percentage changes 
are displayed in the following dynamic plot. 


```{r echo = FALSE}
percen <- function(x){
  x2 <-as.numeric(x)
  out <- (x2[-1] - x2[-length(x2)])/x2[-length(x2)] * 100
  out <- as.xts(out, order.by = index(x)[-1])
}

dat %>% Cl() %>% percen() %>% dygraph()
```


Auto-correlation plot of the percentage change data 
is given below.


```{r echo = FALSE}
suppressMessages(library(forecast))

dat %>% Cl() %>% percen() %>% ggAcf()
```


It is apparent that percentage change data is 
zero-mean, but not constant variance. 
Moreover, it serially uncorrelated as 
almost all the correlation coefficients 
are within the confidence bands. 
It is also interesting to see evidence 
of seasonality, although not significant, 
in the auto-correlation plot as hinted 
by the consistently positive and negative 
s-shaped correlations. 

Based on these, the so-called stochastic 
volatility model would be appropriate to 
model one-period percentage return data. 
For inference we wil take a Bayesian paradigm, 
and sample from the posterior distributions; 
details are skipped.

In- and out-of-sample (20 days; 23 December 2019 - 15 January 2020) 
predictions are displayed in the following plot, 
where while the former predictions are in solid 
lines, the latter in dashed-lines. Note that 
the y-axis is hundred times the variance. 

```{r echo = FALSE}
suppressMessages(library(stochvol))

x <- dat %>% Cl() %>% percen() %>% as.numeric() + 0.0003

res <- svsample(x, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = 0.1)

#summary(res, showlatent = FALSE)
volplot(res, forecast = 20)
```

The posterior means of the out-of-sample 
predictions indicate that volatility tends 
to decrease. However, the distributions are 
highly right-skewed. This indicates that 
high volatilities are possible. 
