---
title: "Robust Joint Models"
author: "Özgür Asar"
date: 2019-07-17
categories: ["R"]
tags: ["joint models", "regression"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Robust Joint Modelling Framework

Joint modelling of longitudinal time-to-event outcomes typically 
combines a linear mixed-effects model for repeated measures 
and a Cox model with time-varying frailty for time-to-event outcome ([Asar et al., 2015](https://academic.oup.com/ije/article/44/1/334/657852)). 
Typical distributional assumption is that random-effects and 
measurement error terms in mixed-effects model are Gaussian. 
However, this assumption might be restricive for real-life 
problems, where it is quite likely to have 

* subjects who do not 
conform the population averaged trends (they are examples of 
outliers in the random-effects), and 

* subjects who has a few 
observations that are quite different compared to the rest 
of the observations for subjects' own collection of measurements 
(they are examples of outliers in measurement error). 

Gaussian distribution would not give appropriate weights to the outliers, 
hence inference might be biased and inefficient, and personalised predictions might be 
misleading. A natural approach would be to replace the Gaussian assumption 
with t distribution. Technical details of joint models with t distributions, 
and associated inferential methods are 
skipped here, and interested reader is referred to [Asar, Fournier and Dantan (2019)](https://arxiv.org/abs/1905.00816). 

## Implementation

We describe the 
R package [`robjm`](https://github.com/ozgurasarstat/robjm) to implement the 
joint models with Gaussian and t distributed random-effects and error terms, 
and subsequently to obtain personalised dynamic predictions. 
For illustration, we will use the 
AIDS data-set (first 250 subjects only). 
Note that the biomarker of interest is the CD4 cell counts, and the 
survival event is death. 

`robjm` is still under development, hence is currently only available from Github. 
Note that `robjm` internatlly calls `tidyverse` and `rstan` packages. 

To install `robjm` from Github and load into the working environment, 
use the following lines:
```{r cars}
#devtools::install_github("ozgurasarstat/robjm", quiet = TRUE)
suppressMessages(library(robjm))
```
AIDS data-set can be loaded prepared for analysis using
```{r}
data(aids)
data(aids.id)

aids$drug2 <- ifelse(aids$drug == "ddC", 0, 1)
aids.id$drug2 <- ifelse(aids.id$drug == "ddC", 0, 1)

idlist <- aids.id$patient

long_data <- aids[aids$patient %in% idlist[1:250], ]
surv_data <- aids.id[aids.id$patient %in% idlist[1:250], ]
```
Below, we first fit the joint model with Gaussian random effects and 
Gaussian error terms, and then t distributed random effects and 
t distributed error terms.
```{r}
## normal normal model
fit_nor_nor <- fit_jm(fixed_long = CD4 ~ obstime, 
                      random_long = ~ obstime, 
                      fixed_surv = cbind(Time, death) ~ drug2, 
                      data_long = long_data,
                      data_surv = surv_data,
                      id_long = "patient",
                      id_surv = "patient",
                      model = "nor_nor",
                      timeVar = "obstime",
                      bh = "weibull",
                      chains = 4,
                      cores = 4,
                      iter = 200,
                      warmup = 100,
                      control = list(adapt_delta = 0.9)
                      )

fit_t_t <- fit_jm(fixed_long = CD4 ~ obstime, 
                  random_long = ~ obstime, 
                  fixed_surv = cbind(Time, death) ~ drug2, 
                  data_long = long_data,
                  data_surv = surv_data,
                  id_long = "patient",
                  id_surv = "patient",
                  model = "t_t_mod3",
                  timeVar = "obstime",
                  bh = "weibull",
                  chains = 4,
                  cores = 4,
                  iter = 200,
                  warmup = 100,
                  control = list(adapt_delta = 0.9)
                  )
```

```{r}
print(fit_nor_nor$res, 
      pars = c("alpha", "Sigma", "sigmasq", "log_lambda", 
               "log_nu", "omega", "eta"))
print(fit_t_t$res, 
      pars = c("alpha", "Sigma", "sigmasq", "phi", "delta", 
               "log_lambda", "log_nu", "omega", "eta"))
```
Traceplots of the MCMC samples can be visualised by
```{r}
traceplot(fit_nor_nor$res, 
          pars = c("alpha", "Sigma", "sigmasq", "log_lambda", 
                   "log_nu", "omega", "eta"),
          inc_warmup = FALSE)
traceplot(fit_t_t$res, 
          pars = c("alpha", "Sigma", "sigmasq", "phi", "delta", 
                   "log_lambda", "log_nu", "omega", "eta"),
          inc_warmup = FALSE)
```
Dynamic predictions for a new subject, say 251th subject, can be obtained by
```{r message = FALSE}
newdata <- dplyr::filter(aids, patient == idlist[251])

fore_nor_nor <- predSurv_jm(object = fit_nor_nor, 
                            newdata = newdata, 
                            forecast = list(h = 5, n = 5),
                            B_control = list(nsel_b = 1, init = 0)
                            )
 
fore_t_t <- predSurv_jm(object = fit_t_t, 
                        newdata = newdata, 
                        forecast = list(h = 5, n = 5),
                        B_control = list(nsel_b = 1, init = 0)
                        ) 
```
The output would be displayed as a matrix by
```{r}
fore_nor_nor$output
fore_t_t$output
```
