---
title: "Robust Joint Models"
author: "Özgür Asar"
date: 2019-07-17
categories: ["R"]
tags: ["joint models", "regression"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Robust Joint Modelling Framework

Joint modelling of longitudinal time-to-event outcomes typically 
combines a linear mixed-effects model and a Cox model [Asar et al.(2015)](https://academic.oup.com/ije/article/44/1/334/657852). 
Typical distributional assumption is that random-effects and 
measurement error term in mixed-effects model are Gaussian. 
However, this assumption might not be optimal for real-life 
problems where it is quite likely to have subjects who do not 
conform the population averaged trends (they are examples of 
outliers in the random-effects), and subjects who has a few 
observations that are quite different compared to the rest 
of the observations for subjects' own collection of measurements 
(they are examples of outliers in the measurement error).  
Gaussian distribution woud not give appropriate weights to the outliers, 
hence inference might be biased and inefficient, and predictions might be 
misleading. A natural approach would be to replace the Gaussian assumption 
with t distribution. Technical details of non-Gaussian joint modelling is 
skipped here, an interested reader is referred to [Asar, Fournier and Dantan](https://arxiv.org/abs/1905.00816). Below, we describe the 
R package [robjm](https://github.com/ozgurasarstat/robjm) to fit the 
joint models with Gaussian and t distributed random-effects and subsequently 
to obtain dynamic predictions. 

## Implementation

Present code to fit the robust models

```{r cars}
library(MASS)
```

